---
output: 
  github_document:
    df_print: kable
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Travis-CI Build Status](https://travis-ci.org/paleolimbot/tidypaleo.svg?branch=master)](https://travis-ci.org/paleolimbot/tidypaleo) [![Coverage Status](https://img.shields.io/codecov/c/github/paleolimbot/tidypaleo/master.svg)](https://codecov.io/github/paleolimbot/tidypaleo?branch=master)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  dpi = 300
)
```

# tidypaleo

Working with multi-proxy paleo-archive data is inherently difficult. There are multiple locations, multiple parameters, and a lot of discipline-specific norms for plot layout and notation. This package simplifies a few workflows to promote the use of R and reproducible documents in paleo-based studies.

## Installation

You can install tidypaleo from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("paleolimbot/tidypaleo")
```

To get the most out of strat diagrams, you will also need the [patchwork package](https://github.com/thomasp85/patchwork) by [\@thomasp85]((https://github.com/thomasp85). You can also install this from GitHub:

```{r gh-installation-patchwork, eval = FALSE}
# install.packages("devtools")
devtools::install_github("thomasp85/patchwork")
```

## Examples

### Age-depth models

### Stratigraphic diagrams

This package exposes a number of functions useful when creating stratigraphic diatrams, including `facet_abundanceh()`, which combines several other functions to help create stratigraphic plots using **ggplot2**. The `geom_col_segsh()` geometry draws horizontal sements, which are commonly used to show species abundance data.

```{r keji-strat}
library(ggplot2)
library(tidypaleo)
theme_set(theme_bw())

ggplot(keji_lakes_plottable, aes(x = rel_abund, y = depth)) +
  geom_col_segsh() +
  scale_y_reverse() +
  facet_abundanceh(vars(taxon), grouping = vars(location)) +
  labs(y = "Depth (cm)")
```

Relative abundance data with a discrete y axis can be drawn using `geom_colh()` and `position_dodgev()` from the **ggstance** package.

```{r halifax-lakes-plot}
ggplot(halifax_lakes_plottable, aes(x = rel_abund, y = location, fill = sample_type)) +
  geom_colh(width = 0.5, position = "dodgev") +
  facet_abundanceh(vars(taxon)) +
  labs(fill = NULL, y = NULL)
```

Age depth models can be constructed (using `age_depth_model()`) and used to draw second axes on stratigraphic plots (using `scale_y_depth_age()`). The `facet_geochem()` function wraps `facet_wrap()`, and includes a labeller (`label_geochem`) that can add units, and automatically renames `d15N`, `d13C` and `d18O` such that they are labelled correctly.

```{r alta-strat}
alta_adm <- age_depth_model(
  alta_lake_bacon_ages, 
  depth = depth_cm,
  age = 1950 - age_weighted_mean_year_BP
)

ggplot(alta_lake_geochem, aes(x = value, y = depth)) +
  geom_lineh() +
  geom_point() +
  scale_y_depth_age(alta_adm, age_name = "Age (Year AD)") +
  facet_geochem_wrap(
    vars(param), 
    nrow = 1, 
    units = c(
      C = "%",
      Cu = "ppm",
      Ti = "ppm",
      d15N = "‰",
      d13C = "‰"
    )
  ) +
  labs(y = "Depth (cm)", x = NULL)
```
