---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# tidypaleo

<!-- badges: start -->
![R-CMD-check](https://github.com/paleolimbot/tidypaleo/workflows/R-CMD-check/badge.svg)
[![Coverage Status](https://img.shields.io/codecov/c/github/paleolimbot/tidypaleo/master.svg)](https://codecov.io/github/paleolimbot/tidypaleo?branch=master)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

Working with multi-proxy paleo-archive data can be difficult. There are multiple locations, multiple parameters, and a lot of discipline-specific norms for plot layout and notation. This package simplifies a few workflows to promote the use of R for reproducible documents in paleo-based studies.

## Installation

You can install tidypaleo from github with:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("paleolimbot/tidypaleo")
```

## Examples

### Strat diagrams

This package exposes a number of functions useful when creating stratigraphic diagrams, including `facet_abundanceh()`, which combines several other functions to help create stratigraphic plots using **ggplot2**. The `geom_col_segsh()` geometry draws horizontal segments, which are commonly used to show species abundance data.

```{r keji-strat, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidypaleo)
theme_set(theme_paleo())

ggplot(keji_lakes_plottable, aes(x = rel_abund, y = depth)) +
  geom_col_segsh() +
  scale_y_reverse() +
  facet_abundanceh(vars(taxon), grouping = vars(location)) +
  labs(y = "Depth (cm)")
```
